---
title: "CMap Data Analysis Workflow"
author: "Hengxin Pan"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{CMap Data Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# CMap Data Analysis Workflow

This vignette provides a detailed walkthrough of the CMap data analysis workflow using the CONCERTDR package. We'll cover the process from extracting parameters from the CMap database to processing combinations and generating the final data extracts.

## Prerequisites

Before starting this workflow, you should have access to the following CMap data files:

- `siginfo_beta.txt`: Contains information about all signatures in the database
- `geneinfo_beta.txt`: Contains information about all genes in the database
- `level5_beta_trt_cp_n720216x12328.gctx`: Contains the expression data for all signatures

These files can be downloaded from the [Broad Institute's CMap website](https://clue.io/) after registration.

```{r setup}
library(CONCERTDR)
```

## Step 1: Extract Parameters from CMap Data

The first step is to analyze the CMap `siginfo_beta.txt` file to extract all available experimental parameters (time points, dosages, cell lines) and save them to a configuration file:

```{r extract_params}
# Adjust the file path to your siginfo_beta.txt file
params <- extract_cmap_parameters(
  siginfo_file = "path/to/siginfo_beta.txt",
  config_dir = "conf",
  verbose = TRUE
)
```

This function:
1. Reads the siginfo file
2. Extracts unique time points, dosages, and cell lines
3. Creates a configuration file (by default at `conf/cmap_options.conf`)

Let's examine the extracted parameters:

```{r examine_params}
# Number of time points, doses, and cell lines
cat("Time points:", length(params$times), "\n")
cat("Dosages:", length(params$doses), "\n")
cat("Cell lines:", length(params$cells), "\n")

# View the first few of each
head(params$times)
head(params$doses)
head(params$cells)
```

## Step 2: Edit the Configuration File

The configuration file contains all available parameters and a `[selected]` section at the bottom where you can specify which parameters to use:

```
[selected]
times = 1,3,5  # Selects the 1st, 3rd, and 5th time points
doses = 1      # Selects only the first dose
cells = all    # Selects all cell lines
```

You can edit this file manually to select the specific parameters you want to use. For this vignette, we'll create a template and then edit it programmatically:

```{r edit_config}
# Create a template config file
config_file <- create_cmap_config_template(
  dest_dir = "conf",
  template_name = "my_config.conf",
  overwrite = TRUE
)

# Read the template
config_lines <- readLines(config_file)

# Find the [selected] section
selected_idx <- which(grepl("\\[selected\\]", config_lines))

# Update selections (for example, selecting the first time, dose, and cell)
# This would normally be done by manually editing the file
if (length(selected_idx) > 0) {
  # Find where each parameter is defined in the selected section
  times_idx <- grep("^times\\s*=", config_lines[selected_idx:length(config_lines)])
  doses_idx <- grep("^doses\\s*=", config_lines[selected_idx:length(config_lines)])
  cells_idx <- grep("^cells\\s*=", config_lines[selected_idx:length(config_lines)])
  
  # Update the selections (adding the offset from selected_idx)
  if (length(times_idx) > 0)
    config_lines[selected_idx + times_idx[1] - 1] <- "times = 1,2"
  if (length(doses_idx) > 0)
    config_lines[selected_idx + doses_idx[1] - 1] <- "doses = 1"
  if (length(cells_idx) > 0)
    config_lines[selected_idx + cells_idx[1] - 1] <- "cells = 1,3"
  
  # Write the updated config back
  writeLines(config_lines, config_file)
}
```

## Step 3: Generate Combinations

Once you've edited the configuration file, generate the combinations of parameters:

```{r generate_combinations}
combinations <- generate_combinations_from_config(config_file)

# Examine the combinations
head(combinations)

# Total number of combinations
nrow(combinations)
```

Each row in the combinations data frame represents a specific experimental condition (time point, dosage, cell line) for which we want to extract data.

## Step 4: Process Combinations and Extract Data

Now we can process these combinations to extract the corresponding data from the GCTX file:

```{r process_combinations}
# This step requires the actual CMap data files
output_files <- process_combinations(
  combinations,
  output_dir = "output",
  geneinfo_file = "path/to/geneinfo_beta.txt",
  siginfo_file = "path/to/siginfo_beta.txt",
  gctx_file = "path/to/level5_beta_trt_cp_n720216x12328.gctx"
)

# View the output files
head(output_files)
```

Alternatively, you can extract all combinations at once into a single data frame:

```{r extract_all}
combined_data <- extract_cmap_data_from_config(
  config_file = config_file,
  geneinfo_file = "path/to/geneinfo_beta.txt",
  siginfo_file = "path/to/siginfo_beta.txt",
  gctx_file = "path/to/level5_beta_trt_cp_n720216x12328.gctx"
)

# Check the dimensions of the extracted data
dim(combined_data)

# View the first few rows and columns
combined_data[1:5, 1:5]

# Retrieve the metadata (time, dose, cell information for each column)
metadata <- attr(combined_data, "metadata")
head(metadata)
```

## Step A5: High Performance Computing with SLURM

For large datasets, you might want to use a high-performance computing cluster with SLURM. You can generate a SLURM submission script:

```{r slurm_script}
# Save combinations to a file
combinations_file <- "combinations.txt"
write.table(combinations, file = combinations_file, row.names = FALSE)

# Generate a SLURM script
slurm_script <- generate_slurm_script(
  combinations_file = combinations_file,
  output_file = "run_cmap_extract.sh",
  job_name = "cmap_extract",
  wall_time = "02:00:00",
  memory = "16G",
  cpus_per_task = 2
)

# The script can then be submitted with:
# sbatch run_cmap_extract.sh
```

## Step 6: Interactive Workflow

For a more interactive experience, you can use the interactive workflow function:

```{r interactive, eval=FALSE}
# This launches an interactive command-line interface
interactive_cmap_workflow()
```

This will guide you through the process with prompts for each step.

## Conclusion

This vignette has walked through the essential steps for extracting data from the CMap database using the CONCERTDR package. The extracted data can then be used for signature matching, as described in the [Signature Matching vignette](signature_matching.html).

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
