---
title: "Introduction to CONCERTDR"
author: "Hengxin Pan"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to CONCERTDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# CONCERTDR: Drug Response Data Analysis Tools

## Overview

CONCERTDR is an R package that provides a comprehensive suite of tools for drug response data analysis, with a particular focus on working with the Connectivity Map (CMap) database. It enables researchers to efficiently extract, process, and match drug response signatures from their experimental data with reference profiles from CMap.

## Installation

To install the latest version of CONCERTDR from Bioconductor:

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("CONCERTDR")
```

For the development version:

```{r installation-dev, eval=FALSE}
# Install devtools if not already installed
# install.packages("devtools")
devtools::install_github("zeratulhx/DR_pipeline")
```

## Dependencies

CONCERTDR depends on the following packages:

- **BiocManager**: For installing Bioconductor packages
- **cmapR**: For working with CMap GCTX files
- **data.table**: For efficient data manipulation
- **RCSM**: For signature matching methods (installed automatically from GitHub)

## Package Structure

The CONCERTDR package is organized into several main components:

1. **CMap Parameter Management and Data Extraction**
   - Functions for extracting experimental parameters and creating configurations
   - Tools for generating combinations of time points, doses, and cell lines
   - GCTX file data extraction functions

2. **Signature Matching**
   - Multiple methods for matching drug response signatures
   - Statistical analysis of matching results
   - Result visualization and interpretation

3. **Utilities**
   - Configuration file management
   - Interactive selection of parameters
   - Batch processing with SLURM support

## Getting Started

See the other vignettes for detailed workflows:

- [CMap Workflow](cmap_workflow.html): Step-by-step guide to extracting and processing CMap data
- [Signature Matching](signature_matching.html): How to match drug response signatures

## Acquiring CMap Data

The Connectivity Map (CMap) dataset can be downloaded from the [Broad Institute's CMap website](https://clue.io/). You'll need to register for an account to access the data.

Required files include:
- `siginfo_beta.txt`: Contains information about all signatures in the database
- `geneinfo_beta.txt`: Contains information about all genes in the database
- `level5_beta_trt_cp_n720216x12328.gctx`: Contains the expression data for all signatures

## Example Workflow

Here's a brief overview of a typical workflow:

```{r example, eval=FALSE}
library(CONCERTDR)

# Create a configuration template
create_cmap_config_template()

# Edit the configuration file manually...
# Then generate combinations
combinations <- generate_combinations_from_config("conf/cmap_options_template.conf")

# Process combinations to extract data
process_combinations(combinations,
                    output_dir = "output",
                    geneinfo_file = "path/to/geneinfo_beta.txt",
                    siginfo_file = "path/to/siginfo_beta.txt",
                    gctx_file = "path/to/level5_beta_trt_cp_n720216x12328.gctx")

# Match a signature against the reference profiles
process_signature_with_df(
  signature_file = "path/to/signature.txt", 
  reference_df = reference_data,
  output_dir = "results",
  methods = c("ks", "xcos") # Example using just two methods
)
```

## Session Info

```{r session-info, eval=TRUE}
sessionInfo()
```
